The Multi-Stencil Framework: A Component-Based Approach to Orchestrate Stencils

### 2017-03-06 ARTICLES/IJPP/articleR1/article.tex
corrections abstract + intro

%----------------------------------------
\title{The Multi-Stencil Framework: A Component-Based Approach to Orchestrate Stencils}
%----------------------------------------
\author{H\'el\`ene Coullon \and Julien Bigot \and Christian Perez}
%----------------------------------------
\institute{H\'el\`ene Coullon \at
              DAPI IMT Atlantique, LS2N, Inria. Nantes, France\\
              \email{helene.coullon@inria.fr}
	\and
           Julien Bigot \at
           Maison de la Simulation, CEA, CNRS, Univ. Paris-Sud, UVSQ, Universit\'e Paris-Saclay, 91191 Gif-sur-Yvette, France\\
           \email{julien.bigot@cea.fr}           
           \and
	Christian Perez \at
              Univ. Lyon, Inria, CNRS, ENS de Lyon. Lyon, France\\
              \email{christian.perez@inria.fr}
}
%----------------------------------------
\maketitle
%----------------------------------------
\begin{abstract}
As the computation power of modern high performance architectures increases, their heterogeneity and complexity also become more important. One of the big challenges of exascale is to reach programming models that give access to high performance computing (HPC) to many scientists and not only to a few HPC specialists. One relevant solution to ease parallel programming for scientists is Domain Specific Language (DSL). However, one problem to avoid with DSLs is to mutualize existing codes and libraries instead of implemented from scratch. For example, this phenomenon occurs for stencil-based numerical simulations, for which a large number of languages has been proposed without code reuse between them. 
The Multi-Stencil Framework (MSF) presented in this paper combines a meta-model to component-based programming models to enhance code reuse and separation of concerns in the specific case of stencils. MSF can easily choose one parallelization technique or another as well as one optimization or another, as well as one back-end implementation or another. It is shown that MSF can reach same performances than a non-component-based MPI implementation over 16.384 cores. Finally, the performance model of the framework for hybrid parallelization is validated by evaluations.

>>>>

%----------------------------------------
\title{The Multi-Stencil Framework: A Component-Based Approach to Orchestrate Stencils}
%----------------------------------------
\author{H\'el\`ene Coullon \and Julien Bigot \and Christian Perez}
%----------------------------------------
\institute{H\'el\`ene Coullon \at
              DAPI IMT Atlantique, LS2N, Inria. Nantes, France\\
              \email{helene.coullon@inria.fr}
	\and
           Julien Bigot \at
           Maison de la Simulation, CEA, CNRS, Univ. Paris-Sud, UVSQ, Universit\'e Paris-Saclay, 91191 Gif-sur-Yvette, France\\
           \email{julien.bigot@cea.fr}           
           \and
	Christian Perez \at
              Univ. Lyon, Inria, CNRS, ENS de Lyon. Lyon, France\\
              \email{christian.perez@inria.fr}
}
%----------------------------------------
\maketitle
%----------------------------------------
\begin{abstract}
As the computation power of modern high performance architectures increases, their heterogeneity and complexity also become more important. One of the big challenges of exascale is to reach programming models that give access to high performance computing (HPC) to many scientists and not only to a few HPC specialists. One relevant solution to ease parallel programming for scientists is Domain Specific Language (DSL). However, one problem to avoid with DSLs is to mutualize existing codes and libraries instead of implemented each solution from scratch. For example, this phenomenon occurs for stencil-based numerical simulations, for which a large number of languages has been proposed without code reuse between them. 
The Multi-Stencil Framework (MSF) presented in this paper combines a new DSL to component-based programming models to enhance code reuse and separation of concerns in the specific case of stencils. MSF can easily choose one parallelization technique or another, one optimization or another, as well as one back-end implementation or another. It is shown that MSF can reach same performances than a non component-based MPI implementation over 16.384 cores. Finally, the performance model of the framework for hybrid parallelization is validated by evaluations.

### 2017-03-06 ARTICLES/IJPP/articleR1/introduction.tex
corrections abstract + intro

One of the big challenges to solve to be able to use these upcoming Exascale computers is to propose programming models which gives access to high performance computing (HPC) to many scientists and not only to a few HPC specialists~\cite{ETP4HPC2013}. Actually, applications that run on supercomputers and need such computation power (\textit{e.g.} physics, weather or genomic) are typically not implemented by HPC specialists but by domain scientists.

%JB: pourquoi tu parles de taches ?
One possible runtime execution model for HPC is based on dynamic scheduling of task graphs combined with message passing~\cite{Gautier:2013:XRS:2510661.2511383,Augonnet2011,wu:hal-01078359}.
These models increase code portability and enable users to reach interesting performance on heterogeneous architectures.
%JB: j'enleverais ce qui est au dessus et je commencerait tout de suite avec "OpenMP c'est compliqué"
General purpose parallel languages, such as OpenMP~\cite{660313} and OpenCL~\cite{Stone:2010:OPP:622179.1803953} start to include these concepts.
For non-expert users however, these languages are difficult to use and make it complex to tune applications for a given architecture. 
An interesting approach that combines simplicity of use due to a high abstraction level with efficient execution consists in domain specific languages (DSL) and domain specific frameworks (DSF).
These solutions are specific to a given domain and propose a grammar or an API which is easy to understand for specialists of this domain.
Moreover, knowledge about the targeted domain can be embedded in the compiler that can thus automatically apply parallelization and optimizations techniques to produce high performance code.
Domain specific solutions are therefore able to separate end-user concerns from HPC concerns which is a requirement to make HPC accessible to a wider audience.

Many domain specific languages and frameworks have been proposed that claim to handle distinct specific use-cases each.
Each solution is however typically re-implemented from scratch.
In this paper, we claim that the sharing of common building blocks when designing DSLs or DSFs would increases re-use, flexibility and maintainability in their implementation.
%JB: J'ajoute cet argument, tu peux le retirer si tu trouves ça dangereux
It would also ease the creation of approaches and applications combining multiple DSLs and DSFs.

@@

We propose the Multi-Stencil Framework (MSF) that is built upon a meta-model of multi-stencil simulations. MSF produces a parallel orchestration of a multi-stencil program without being aware of the underlying implementation choices (\eg distributed data structures, task scheduler etc.). Thanks to this meta-model MSF is able to easily switch from one parallelization technique to another and from one optimization to another. Moreover, as the meta-model is independent from implementation details, MSF can easily choose one back-end or another, thus easing code reuse of existing solutions.
To ease composition of existing solutions, MSF is based on component-based programming~\cite{Szyperski:2002:CSB:515228}, where applications are defined as an assembly of building blocks, or components.

After a short overview of the Multi-Stencil Framework given in Section~\ref{sect:msf}, the paper is organized as follows. A meta-model of a multi-stencil program is presented in Section~\ref{sect:formalism}; from this meta-model are built both a light and descriptive domain specific language, namely MSL, as well as a generic component assembly of the application described in Section~\ref{sect:msl}; the compiler MSC of the framework is described in Section~\ref{sect:parallelism}; finally a performance evaluation is detailed in Section~\ref{sect:eval} .

>>>>

One of the big challenges to be able to use those upcoming Exascale computers is to propose programming models that give access to high performance computing (HPC) to many scientists and not only to a few HPC specialists~\cite{ETP4HPC2013}. Actually, applications that run on supercomputers and need such computation power (\textit{e.g.} physics, weather or genomic) are typically not implemented by HPC specialists but by domain scientists.

%\JB{pourquoi tu parles de taches ?}
%One possible runtime execution model for HPC is based on dynamic scheduling of task graphs combined with message passing~\cite{Gautier:2013:XRS:2510661.2511383,Augonnet2011,wu:hal-01078359}.
%These models increase code portability and enable users to reach interesting performance on heterogeneous architectures.
%\JB{j'enleverais ce qui est au dessus et je commencerait tout de suite avec "OpenMP c'est compliqué"}
%General purpose parallel languages, such as OpenMP~\cite{660313} and OpenCL~\cite{Stone:2010:OPP:622179.1803953} start to include these concepts.
Many general purpose languages and frameworks have improved the simplicity of writing parallel codes. For example PGAS models~\cite{Nieplocha:2006:AAP:1125980.1125985} or task-based frameworks, such as OpenMP~\cite{660313}, Legion~\cite{bauer:legion:sc:2012} or StarPU~\cite{Augonnet2011}, partially hide intricate details of parallelism to the user. For non-expert users however, these languages and frameworks are still difficult to use. Moreover, tuning an application for a given architecture is still very complex to achieve with these solutions.
An interesting approach that combines simplicity of use, due to a high abstraction level, with efficient execution are domain specific languages (DSL) and domain specific frameworks (DSF).
These solutions are specific to a given domain and propose a grammar or an API which is easy to understand for specialists of this domain.
Moreover, knowledge about the targeted domain can be embedded in the compiler that can thus automatically apply parallelization and optimization techniques to produce high performance code.
Domain specific solutions are therefore able to separate end-user concerns from HPC concerns which is a requirement to make HPC accessible to a wider audience.

Many domain specific languages and frameworks have been proposed. Each one claims to handle a distinct specific optimization or use case.
Each solution is however typically re-implemented from scratch.
In this paper, we claim that the sharing of common building blocks when designing DSLs or DSFs would increases re-use, flexibility and maintainability in their implementation.
It would also ease the creation of approaches and applications combining multiple DSLs and DSFs.

@@

We propose the Multi-Stencil Framework (MSF) that is built upon a meta-formalism of multi-stencil simulations. MSF produces a parallel orchestration of a multi-stencil program without being aware of the underlying implementation choices (\eg distributed data structures, task scheduler etc.). Thanks to this meta-formalism MSF is able to easily switch from one parallelization technique to another and from one optimization to another. Moreover, as MSF is independent from implementation details, MSF can easily choose one back-end or another, thus easing code reuse of existing solutions.
To ease composition of existing solutions, MSF is based on component-based programming~\cite{Szyperski:2002:CSB:515228}, where applications are defined as an assembly of building blocks, or components.

After a short overview of the Multi-Stencil Framework given in Section~\ref{sect:msf}, the paper is organized as follows. The meta-formalism of a multi-stencil program is presented in Section~\ref{sect:formalism}; from this formalism are built both a light and descriptive domain specific language, namely MSL, as well as a generic component assembly of the application described in Section~\ref{sect:msl}; the compiler MSC of the framework is described in Section~\ref{sect:parallelism}; finally a performance evaluation is detailed in Section~\ref{sect:eval} .
